Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-03-01 paper/zhang-cook-laa-langrene-menendez.Rmd
SA -> CRS, SAJO -> SA

```{r external, include = FALSE, cache = FALSE}
library(knitr)
library(tidyverse)
library(ferrn)
read_chunk(here::here('scripts/toy.R'))
read_chunk(here::here('scripts/diagnose.R'))
read_chunk(here::here('scripts/kol_result.R'))
read_chunk(here::here('scripts/implementation.R'))
```

@@

Three optimisers have been implemented in the \CRANpkg{tourr} [@tourr] package: simulated annealing (SA), simulated annealing with jump out (SAJO), and pseudo derivative (PD). Simulated annealing (SA) is a random search optimiser that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is generated randomly. $\mathbf{A}_{l}$ is then orthonormalised to fulfil the basis constraint. If $\mathbf{A}_{l}$ has an index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the optimiser outputs $\mathbf{A}_{l}$ for guided tour to construct an interpolation path. The neighbourhood parameter $\alpha$ is adjusted by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$ before the next iteration starts. The optimiser terminates when the maximum number of iteration $l_{\max}$ is reached before a better basis can be found. The algorithm of SA is summarised in Algorithm \ref{random-search}. @posse1995projection has proposed a slightly different cooling scheme by introducing a halving parameter $c$. In his proposal, the $\alpha$ is only adjusted if the last iteration takes more than $c$ times to find a better basis.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{input}{input}
  \SetKwInOut{output}{output}
    \input{$f(.)$, $\alpha_1$, $l_{\max}$, $\text{cooling}$} 
    \output{$\mathbf{A}_{l}$}
    Generate random start $\mathbf{A}_1$ and set $\mathbf{A}_{\text{cur}} \coloneqq \mathbf{A}_1$, $I_{\text{cur}} = f(\mathbf{A}_{\text{cur}})$, $j = 1$\;
  \Repeat{$\mathbf{A}_l$ is too close to $\mathbf{A}_{\text{cur}}$ in terms of geodesic distance}{
   Set $l = 1$\;
  \Repeat{$l > l_{\max}$ or $I_{l} > I_{\text{cur}}$}{
    Generate $\mathbf{A}_{l} = (1- \alpha_j)\mathbf{A}_{\text{cur}} + \alpha_j \mathbf{A}_{\text{rand}}$ and orthogonalise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    Update $l = l + 1$\;
  }
  Update $\alpha_{j+1} = \alpha_j * \text{cooling}$\;
  Construct the geodesic interpolation between $\mathbf{A}_{\text{cur}}$ and $\mathbf{A}_l$\; 
  Update $\mathbf{A}_{\text{cur}} = \mathbf{A}_l$ and $j = j + 1$\;
}
  \caption{Simulated annealing (SA)}
  \label{random-search}
\end{algorithm}

Simulated annealing with jump out (SAJO) [@kirkpatrick1983optimization; @bertsimas1993simulated] uses the same sampling process as SA but allows a probabilistic acceptance of a basis with lower index value than the current. Given an initial value of $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, SAJO gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the optimiser to make a move and explore the basis space even the landing basis does have a higher index value and hence enable the optimiser to jump out of a local optimum. The algorithm \ref{simulated-annealing} highlights how SAJO differs from SA in the inner loop.

\begin{algorithm}
\SetAlgoLined
\Repeat{$l > l_{\max}$ or $I_{l} > I_{\text{cur}}$ or $P > U$}{
    Generate $\mathbf{A}_{l} = (1- \alpha_j)\mathbf{A}_{\text{cur}} + \alpha_j \mathbf{A}_{\text{rand}}$ and orthogonalise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$, $T(l) = \frac{T_0}{\log(l + 1)}$ and $P= \min\left\{\exp\left[-\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
    Draw $U$ from a uniform distribution: $U \sim \text{Unif(0, 1)}$\;
    Update $l = l + 1$\;
  }
  \caption{Simulated annealing with jump out (SAJO)}
  \label{simulated-annealing}
\end{algorithm}

@@

Three main pieces of information are recorded for the projection pursuit optimisers: 1) projection bases $\mathbf{A}$, 2) index values $I$, and 3) state $S$. For SA and SAJO, possible states include `random_search`,  `new_basis`, and `interpolation`. Pseudo derivative (PD) has a wider variety of states including `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`. Multiple iterators index the information collected at different levels: $t$ is a unique identifier prescribing the natural ordering of each observation;  $j$ and $l$ are the counter of the outer and inner loop respectively. Other parameters of interest recorded include `method` that tags the name of the optimiser, and `alpha` that indicates the sampling neighbourhood size for searching observations.  A matrix notation of the data structure is presented in equation \ref{eq:data-structure}.

@@

Figure \ref{fig:toy-search} shows an example of the search plot for SA (left) and SAJO (right). Both optimisers quickly find better bases in the first few iterations and then take longer to find one in the later iterations. The anchor bases, the ones found with the highest index value in each iteration, always have an increased index value in the optimiser SA while this is not the case for SAJO. This feature gives SA an advantage in this simple example to quickly find the optimum.

```{r toy-search, fig.cap="A comparison of the searches by two optimisers: SA (left) and SAJO (right) on a 2D projection problem of a six-variable dataset, \\code{boa6} using holes index. Both optimisers reach the final basis with a similar index value while it takes SAJO longer to find the final basis."}
```

@@

```{r toy-interp, fig.cap = "An inspection of the index values as the optimisation progress for two optimisers: SA (left) and SAJO (right). The holes index is optimised for a 2D projection problem on the six-variable dataset \\code{boa6}. Lines indicate the interpolation and dots indicate new target bases generated by the optimisers. Interpolation in both optimisation is smooth while SAJO (right) is observed to first pass by some bases with higher index value before reach the target bases in Time 76-130."}
```

Another interesting diagnostic is to examine how the index value improves between interpolating bases since the projection on these bases is played by the tour animation. Trace plots are created by plotting the index value against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisers as Figure \ref{fig:toy-search} and one can observe that the trace is smooth in both cases. It may seem bizarre at the first sight that the interpolation in the left panel first passes some bases with higher index value before decreases to a lower target. This happens because on the one hand, the probabilistic acceptance in SAJO implies that some worse bases will be accepted by the optimiser. On the other hand, guided tour interpolate between the current and target basis to provide a smooth transition between projections. These two facts indicate that a non-monotonic interpolation can't be avoided for SAJO. Later, in section \ref{monotonic}, there will be a discussion on improving the non-monotonic interpolation for SA.

@@

```{r toy-pca, fig.cap = "Search paths of SA (green) and PD (brow) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \\code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. All the bases in PD has been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases."}
```

@@

Figure \ref{fig:toy-pca} shows the PCA plot of SA and PD for a 1D projection problem. Both optimisers find the optimum but PD gets closer. With PCA plot, one can visually appreciate the nature of these two optimisers: PD first evaluates points in a small neighbourhood for a promising direction, while SA evaluates points randomly in the search space to search for the next target. There are dashed lines annotated for SA and it describes the interruption of the interpolation, which will be discussed in details in Section \ref{monotonic}.

@@

```{r toy-pca-animated, fig.cap = "Six frames selected from the animated version of the previous plot. With animation, the progression of the search paths from start to finish are better identified. SA (green) finishes the optimisation quicker than PD (brown) since there is no further movement for SA in the sixth frame. The full video of the animation can be found at \\url{https://vimeo.com/504242845}."}
```

Animation is another type of display to show how the search progresses from start to finish in the space. An  `animate = TRUE` argument is used to enable the animation in the PCA plot. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that SA finds its end basis quicker than PD since SA finishes its search in the 5th frame while PD is still making more progress.

@@

While the previous few examples have looked at the space of 1D bases in a unit sphere, this section visualises the space of 2D bases. Recall that the columns in a 2D basis are orthogonal to each other, the space of $p \times 2$ bases is a $p$-D torus. If take $p = 3$, one would see a classical 3D torus shape as shown by the grey points in Figure \ref{fig:toy-torus}. The two circles of the torus can be observed to be perpendicular to each other and this can be linked back to the orthogonality of the columns in the bases. Two paths optimised by SA and PD are plotted on top of the torus and coloured in green and brown respectively to match the previous few 1D space plots. The final basis found by PD and SA are shown in a larger shape  and printed below respectively:

@@

```{r toy-torus, fig.cap="Six frames selected from rotating the 2D basis space along with two search paths optimised by PD (brown) and SA (green). The projection problem is a 2D projection with three variables using holes index. The grey points are randomly generated 2D projection bases in the space and it can be observed that these points form a torus. The full video of the animation can be found at \\url{https://vimeo.com/512882784}."}
```

@@

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with higher index value than the target one. For SAJO, a non-monotonic interpolation is justified since target bases do not necessarily have a higher index value than the current one, while this is not the case for SA. The original trace plot for a 2D projection problem, optimised by SA, is shown on the left panel of Figure \ref{fig:interruption} and one can observe clearly that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop at the best basis found in the interpolation. The right panel of Figure \ref{fig:interruption} shows the trace plot after implementing the interruption and while the first two interpolations are identical, the basis at Time 61 has a higher index value than the target in the third interpolation. Rather than starting the next iteration from the target basis on Time 65, SA starts the next iteration at Time 61 on the right panel and reaches a better final basis.

```{r interruption, fig.cap = "Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \\code{boa6} data using holes index, optimised by SA. On the left panel, basis with higher index value is found during the interpolation but not used. On the right panel, the interruption stops the interpolation at the basis with the highest index value for each iteration and results in a final basis with higher index value, as shown on the right panel."}
```

@@

```{r polish, fig.height= 5, fig.cap = "Comparison of the projected data before and after using polishing along with the trace of index value of a 2D projection problem on \\code{boa6} data using holes index. Optimiser SA is used before the polish. The clustering structure in the data is detected by SA but the polish step improves the index value and produces clearer boundaries of the clusters, espeically along the vertial axis."}
```

Once the final basis has been found by an optimiser, one may want to push further in the close neighbourhood to see if an even better basis can be found. A polish search takes the final basis of an optimiser as the start of a new guided tour to search for local breakthrough. The polish algorithm is similar to the SA but with three distinctions: 1) a hundred rather than one candidate bases are generated each time in the inner loop; 2) the neighbourhood size is reduced in the inner loop, rather than in the outer loop; and 3) three more termination conditions have been added to ensure the new basis generated is distinguishable from the current one in terms of the distance in the space,  percentage change in the index value, and neighbourhood size:

@@

Figure \ref{fig:polish} presents the trace plot of a 2D projection, optimised by SA and followed by the polish. The projection by the final basis of each algorithm is also shown. The end basis found by SA reveals the four clusters in the data, but the edges of each cluster are not clean-cut. Polish works with this end basis and further pushes the index value to produce much clear edges of the cluster, especially along the vertical axis.

@@

```{r noisy-better-geo, fig.height = 4, fig.cap = "Comparison of the three optimisers in optimising $I^{nk}(n)$ index for a 1D projection problem on a five-variable dataset, \\code{boa5}. Both SA and SAJO succeeds in the optimisation, PD fails to optimise this non-smooth index. Further, SAJO takes  much longer than SA to finish the optimisation, it finishes off closer to the theoretical best."}
```

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers and as expected, none of the interpolated path is smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index. While SA and SAJO have managed to get close to the index value of the theoretical best, trace plot shows that it takes SAJO much longer to interpolate towards the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SAJO tends to generously accept inferior bases before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. Pseudo Derivative (PD) quickly gets stuck near the starting position. Comparing the amount of random search done by SA and SAJO, it is surprising that SAJO does not carry as many samples as SA. Combining the insights from both the trace and PCA plot, one can learn the two different search strategies by SA and SAJO: SA frequently samples in the space and only make a move when an improvement is guaranteed to be made, while SAJO first broadly accepts bases in the space and then starts the extensive sampling in a narrowed subspace. The specific decision of which optimiser to use will depend on the index curve in the basis space but if the basis space is non-smooth, accepting inferior bases at first, as what SAJO has done, can lead to a more efficient search, in terms of the overall number of points evaluated.

```{r kol-result, fig.height = 4, fig.cap="Comparing 20 search paths in the PCA-projected basis space facetted by two optimisers: SA and SAJO, and two search sizes: 0.5 and 0.7. The optimisation is on the 1D projecton index, $I^{nk}(n)$, for \\code{boa6} data, where a local optimum, annotated by the cross (x), is presented in this experiment, along with the global optimum (*)."}
```

The next experiment compares the performance of SA and SAJO when a local maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared to understand how a large search neighbourhood would affect the final basis and the length of the search. Figure \ref{fig:kol-result} shows 80 paths simulated using 20 seeds in the PCA plot, faceted by  optimiser and search size. With SA and a search size of 0.5, despite being the simplest and fastest, the optimiser fails in three instances where the final basis lands neither near the local nor the global optimum. With a larger search size of 0.7, more seeds have found the global maximum. Comparing SA and SAJO for a search size of 0.5, SAJO does not seem to improve the final basis found, despite having longer interpolation paths. However, the denser paths near the local maximum is an indicator that SAJO is working hard to examine if there is any other optimum in the basis space but the relative small search size has diminished its ability to reach the global maximum. With a larger search size, almost all the seeds (16 out of 20) have found the global maximum and some final bases are much closer to the theoretical best, as compared to the three other cases. This indicates that SAJO, with a reasonable large search size, is able to overcome the local optimum and optimise close towards the global optimum.

@@

```{r flip-sign, fig.cap = "Comparison of the interpolation in the PCA-projected basis space before and after reconciling the orientation of the target basis. Optimisation is on the 1D projection index, $I^{nk}(n)$, for boa6 data using SA with seed 2463. The dots represent the target basis in each iteration and the path shows the interpolation. On the left panel, one target basis is generated with an opposite orientation to the current basis (hence appear on the other side of the basis space) and the interpolator crosses the origin to perform the interpolation. The right panel shows the same interpolation after implementing an orientation check and the undesirable interpolation disappears."}
```

>>>>

```{r external, include = FALSE, cache = FALSE}
library(knitr)
library(tidyverse)
library(ferrn)

knitr::read_chunk(
  here::here("scripts/toy.R")
)

knitr::read_chunk(
  here::here("scripts/diagnose.R")
)

knitr::read_chunk(
  here::here("scripts/implementation.R")
)

knitr::read_chunk(
  here::here("scripts/kol_result.R")
)

```

@@

Three optimisers have been implemented in the \CRANpkg{tourr} [@tourr] package: Creeping random search (CRS), simulated annealing (SA), and pseudo derivative (PD). Creeping random search (CRS) is a random search optimiser that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = (1- \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{\text{rand}}$ is generated randomly. $\mathbf{A}_{l}$ is then orthonormalised to fulfil the basis constraint. If $\mathbf{A}_{l}$ has an index value higher than the current basis $\mathbf{A}_{\text{cur}}$, the optimiser outputs $\mathbf{A}_{l}$ for guided tour to construct an interpolation path. The neighbourhood parameter $\alpha$ is adjusted by a cooling parameter: $\alpha_{j+1} = \alpha_j * \text{cooling}$ before the next iteration starts. The optimiser terminates when the maximum number of iteration $l_{\max}$ is reached before a better basis can be found. The algorithm of SA is summarised in Algorithm \ref{random-search}. @posse1995projection has proposed a slightly different cooling scheme by introducing a halving parameter $c$. In his proposal, the $\alpha$ is only adjusted if the last iteration takes more than $c$ times to find a better basis.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{input}{input}
  \SetKwInOut{output}{output}
    \input{$f(.)$, $\alpha_1$, $l_{\max}$, $\text{cooling}$} 
    \output{$\mathbf{A}_{l}$}
    Generate random start $\mathbf{A}_1$ and set $\mathbf{A}_{\text{cur}} \coloneqq \mathbf{A}_1$, $I_{\text{cur}} = f(\mathbf{A}_{\text{cur}})$, $j = 1$\;
  \Repeat{$\mathbf{A}_l$ is too close to $\mathbf{A}_{\text{cur}}$ in terms of geodesic distance}{
   Set $l = 1$\;
  \Repeat{$l > l_{\max}$ or $I_{l} > I_{\text{cur}}$}{
    Generate $\mathbf{A}_{l} = (1- \alpha_j)\mathbf{A}_{\text{cur}} + \alpha_j \mathbf{A}_{\text{rand}}$ and orthogonalise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    Update $l = l + 1$\;
  }
  Update $\alpha_{j+1} = \alpha_j * \text{cooling}$\;
  Construct the geodesic interpolation between $\mathbf{A}_{\text{cur}}$ and $\mathbf{A}_l$\; 
  Update $\mathbf{A}_{\text{cur}} = \mathbf{A}_l$ and $j = j + 1$\;
}
  \caption{Creeping random search (CRS)}
  \label{random-search}
\end{algorithm}

Simulated annealing (SA) [@kirkpatrick1983optimization; @bertsimas1993simulated] uses the same sampling process as CRS but allows a probabilistic acceptance of a basis with lower index value than the current. Given an initial value of $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. When a candidate basis fails to have an index value larger than the current basis, SAJO gives it a second chance to be accepted with probability $$P= \min\left\{\exp\left[-\frac{\mid I_{\text{cur}} - I_{l} \mid}{T(l)}\right],1\right\}$$ where $I_{(\cdot)}$ denotes the index value of a given basis. This implementation allows the optimiser to make a move and explore the basis space even the landing basis does have a higher index value and hence enable the optimiser to jump out of a local optimum. The algorithm \ref{simulated-annealing} highlights how SAJO differs from SA in the inner loop.

\begin{algorithm}
\SetAlgoLined
\Repeat{$l > l_{\max}$ or $I_{l} > I_{\text{cur}}$ or $P > U$}{
    Generate $\mathbf{A}_{l} = (1- \alpha_j)\mathbf{A}_{\text{cur}} + \alpha_j \mathbf{A}_{\text{rand}}$ and orthogonalise $\mathbf{A}_{l}$\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$, $T(l) = \frac{T_0}{\log(l + 1)}$ and $P= \min\left\{\exp\left[-\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
    Draw $U$ from a uniform distribution: $U \sim \text{Unif(0, 1)}$\;
    Update $l = l + 1$\;
  }
  \caption{Simulated annealing (SA)}
  \label{simulated-annealing}
\end{algorithm}

@@

Three main pieces of information are recorded for the projection pursuit optimisers: 1) projection bases $\mathbf{A}$, 2) index values $I$, and 3) state $S$. For CRS and SA, possible states include `random_search`,  `new_basis`, and `interpolation`. Pseudo derivative (PD) has a wider variety of states including `new_basis`, `direction_search`, `best_direction_search`, `best_line_search`, and `interpolation`. Multiple iterators index the information collected at different levels: $t$ is a unique identifier prescribing the natural ordering of each observation;  $j$ and $l$ are the counter of the outer and inner loop respectively. Other parameters of interest recorded include `method` that tags the name of the optimiser, and `alpha` that indicates the sampling neighbourhood size for searching observations.  A matrix notation of the data structure is presented in equation \ref{eq:data-structure}.

@@

Figure \ref{fig:toy-search} shows an example of the search plot for CRS (left) and SA (right). Both optimisers quickly find better bases in the first few iterations and then take longer to find one in the later iterations. The anchor bases, the ones found with the highest index value in each iteration, always have an increased index value in the optimiser CRS while this is not the case for SA. This feature gives SA an advantage in this simple example to quickly find the optimum.

```{r toy-search, fig.cap="A comparison of the searches by two optimisers: CRS (left) and SA (right) on a 2D projection problem of a six-variable dataset, \\code{boa6} using holes index. Both optimisers reach the final basis with a similar index value while it takes SA longer to find the final basis."}
```

@@

```{r toy-interp, fig.cap = "An inspection of the index values as the optimisation progress for two optimisers: CRS (left) and SA (right). The holes index is optimised for a 2D projection problem on the six-variable dataset \\code{boa6}. Lines indicate the interpolation and dots indicate new target bases generated by the optimisers. Interpolation in both optimisation is smooth while SA is observed to first pass by some bases with higher index value before reach the target bases in Time 76-130."}
```

Another interesting diagnostic is to examine how the index value improves between interpolating bases since the projection on these bases is played by the tour animation. Trace plots are created by plotting the index value against time. Figure \ref{fig:toy-interp} presents the trace plot of the same optimisers as Figure \ref{fig:toy-search} and one can observe that the trace is smooth in both cases. It may seem bizarre at the first sight that the interpolation in the left panel first passes some bases with higher index value before decreases to a lower target. This happens because on the one hand, the probabilistic acceptance in SA implies that some worse bases will be accepted by the optimiser. On the other hand, guided tour interpolate between the current and target basis to provide a smooth transition between projections. These two facts indicate that a non-monotonic interpolation can't be avoided for SA. Later, in section \ref{monotonic}, there will be a discussion on improving the non-monotonic interpolation for CRS.

@@

```{r toy-pca, fig.cap = "Search paths of CRS (green) and PD (brown) in the PCA-reduced basis space for 1D projection problem on the five-variable dataset, \\code{boa5} using holes index. The basis space, a 5D unit sphere, is projected onto a 2D circle by PCA. All the bases in PD has been flipped for easier comparison of the final bases and a grey dashed line has been annotated to indicate the symmetry of the two start bases."}
```

@@

Figure \ref{fig:toy-pca} shows the PCA plot of CRS and PD for a 1D projection problem. Both optimisers find the optimum but PD gets closer. With PCA plot, one can visually appreciate the nature of these two optimisers: PD first evaluates points in a small neighbourhood for a promising direction, while CRS evaluates points randomly in the search space to search for the next target. There are dashed lines annotated for CRS and it describes the interruption of the interpolation, which will be discussed in details in Section \ref{monotonic}.

@@

```{r toy-pca-animated, fig.cap = "Six frames selected from the animated version of the previous plot. With animation, the progression of the search paths from start to finish are better identified. CRS (green) finishes the optimisation quicker than PD (brown) since there is no further movement for CRS in the sixth frame. The full video of the animation can be found at \\url{https://vimeo.com/504242845}."}
```

Animation is another type of display to show how the search progresses from start to finish in the space. An  `animate = TRUE` argument is used to enable the animation in the PCA plot. Figure \ref{fig:toy-pca-animated} shows six frames from the animation of the PCA plot in Figure \ref{fig:toy-pca}. An additional piece of information one can learn from this animation is that CRS finds its end basis quicker than PD since CRS finishes its search in the 5th frame while PD is still making more progress.

@@

While the previous few examples have looked at the space of 1D bases in a unit sphere, this section visualises the space of 2D bases. Recall that the columns in a 2D basis are orthogonal to each other, the space of $p \times 2$ bases is a $p$-D torus. If take $p = 3$, one would see a classical 3D torus shape as shown by the grey points in Figure \ref{fig:toy-torus}. The two circles of the torus can be observed to be perpendicular to each other and this can be linked back to the orthogonality of the columns in the bases. Two paths optimised by CRS and PD are plotted on top of the torus and coloured in green and brown respectively to match the previous few 1D space plots. The final basis found by PD and CRS are shown in a larger shape  and printed below respectively:

@@

```{r toy-torus, fig.cap="Six frames selected from rotating the 2D basis space along with two search paths optimised by PD (brown) and CRS (green). The projection problem is a 2D projection with three variables using holes index. The grey points are randomly generated 2D projection bases in the space and it can be observed that these points form a torus. The full video of the animation can be found at \\url{https://vimeo.com/512882784}."}
```

@@

An example of non-monotonic interpolation has been given in Figure \ref{fig:toy-interp}: a path that passes bases with higher index value than the target one. For SA, a non-monotonic interpolation is justified since target bases do not necessarily have a higher index value than the current one, while this is not the case for CRS. The original trace plot for a 2D projection problem, optimised by CRS, is shown on the left panel of Figure \ref{fig:interruption} and one can observe clearly that the non-monotonic interpolation has undermined the optimiser to realise its full potential. Hence, an interruption is implemented to stop at the best basis found in the interpolation. The right panel of Figure \ref{fig:interruption} shows the trace plot after implementing the interruption and while the first two interpolations are identical, the basis at Time 61 has a higher index value than the target in the third interpolation. Rather than starting the next iteration from the target basis on Time 65, CRS starts the next iteration at Time 61 on the right panel and reaches a better final basis.

```{r interruption, fig.cap = "Comparison of the interpolation before and after implementing the interruption for the 2D projection problem on \\code{boa6} data using holes index, optimised by CRS. On the left panel, basis with higher index value is found during the interpolation but not used. On the right panel, the interruption stops the interpolation at the basis with the highest index value for each iteration and results in a final basis with higher index value, as shown on the right panel."}
```

@@

```{r polish, fig.height= 5, fig.cap = "Comparison of the projected data before and after using polishing along with the trace of index value of a 2D projection problem on \\code{boa6} data using holes index. Optimiser CRS is used before the polish. The clustering structure in the data is detected by CRS but the polish step improves the index value and produces clearer boundaries of the clusters, espeically along the vertial axis."}
```

Once the final basis has been found by an optimiser, one may want to push further in the close neighbourhood to see if an even better basis can be found. A polish search takes the final basis of an optimiser as the start of a new guided tour to search for local breakthrough. The polish algorithm is similar to the CRS but with three distinctions: 1) a hundred rather than one candidate bases are generated each time in the inner loop; 2) the neighbourhood size is reduced in the inner loop, rather than in the outer loop; and 3) three more termination conditions have been added to ensure the new basis generated is distinguishable from the current one in terms of the distance in the space,  percentage change in the index value, and neighbourhood size:

@@

Figure \ref{fig:polish} presents the trace plot of a 2D projection, optimised by CRS and followed by the polish. The projection by the final basis of each algorithm is also shown. The end basis found by CRS reveals the four clusters in the data, but the edges of each cluster are not clean-cut. Polish works with this end basis and further pushes the index value to produce much clear edges of the cluster, especially along the vertical axis.

@@

```{r noisy-better-geo, fig.height = 4, fig.cap = "Comparison of the three optimisers in optimising $I^{nk}(n)$ index for a 1D projection problem on a five-variable dataset, \\code{boa5}. Both CRS and SA succeeds in the optimisation, PD fails to optimise this non-smooth index. Further, SA takes  much longer than CRS to finish the optimisation, it finishes off closer to the theoretical best."}
```

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers and as expected, none of the interpolated path is smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index. While CRS and SA have managed to get close to the index value of the theoretical best, trace plot shows that it takes SA much longer to interpolate towards the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SA tends to generously accept inferior bases before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. Pseudo Derivative (PD) quickly gets stuck near the starting position. Comparing the amount of random search done by CRS and SA, it is surprising that SA does not carry as many samples as CRS. Combining the insights from both the trace and PCA plot, one can learn the two different search strategies by CRS and SA: CRS frequently samples in the space and only make a move when an improvement is guaranteed to be made, while SA first broadly accepts bases in the space and then starts the extensive sampling in a narrowed subspace. The specific decision of which optimiser to use will depend on the index curve in the basis space but if the basis space is non-smooth, accepting inferior bases at first, as what SA has done, can lead to a more efficient search, in terms of the overall number of points evaluated.

```{r kol-result, fig.height = 4, fig.cap="Comparing 20 search paths in the PCA-projected basis space facetted by two optimisers: CRS and SA, and two search sizes: 0.5 and 0.7. The optimisation is on the 1D projecton index, $I^{nk}(n)$, for \\code{boa6} data, where a local optimum, annotated by the cross (x), is presented in this experiment, along with the global optimum (*)."}
```

The next experiment compares the performance of CRS and SA when a local maximum exists. Two search neighbourhood sizes: 0.5 and 0.7 are compared to understand how a large search neighbourhood would affect the final basis and the length of the search. Figure \ref{fig:kol-result} shows 80 paths simulated using 20 seeds in the PCA plot, faceted by  optimiser and search size. With CRS and a search size of 0.5, despite being the simplest and fastest, the optimiser fails in three instances where the final basis lands neither near the local nor the global optimum. With a larger search size of 0.7, more seeds have found the global maximum. Comparing CRS and SA for a search size of 0.5, SA does not seem to improve the final basis found, despite having longer interpolation paths. However, the denser paths near the local maximum is an indicator that SA is working hard to examine if there is any other optimum in the basis space but the relative small search size has diminished its ability to reach the global maximum. With a larger search size, almost all the seeds (16 out of 20) have found the global maximum and some final bases are much closer to the theoretical best, as compared to the three other cases. This indicates that SA, with a reasonable large search size, is able to overcome the local optimum and optimise close towards the global optimum.

@@

```{r flip-sign, fig.cap = "Comparison of the interpolation in the PCA-projected basis space before and after reconciling the orientation of the target basis. Optimisation is on the 1D projection index, $I^{nk}(n)$, for boa6 data using CRS with seed 2463. The dots represent the target basis in each iteration and the path shows the interpolation. On the left panel, one target basis is generated with an opposite orientation to the current basis (hence appear on the other side of the basis space) and the interpolator crosses the origin to perform the interpolation. The right panel shows the same interpolation after implementing an orientation check and the undesirable interpolation disappears."}
```
